\Large \bf Climate Change Awareness and Willingness to Pay for its Mitigation: \\ Evidence from the UK

### 2018-11-15 LiteratureFS/MyReview/Review11.tex
blll

Demand for staple food is usually fixed, even if its production change. Therefore, when staple food becomes scarce in some location, its price tend to increase rapidly. For lower income families, this leads to reduction in calorie intake and often to malnutrition and increased risk of health problems (\citealp{Golden2011, Handa2006}). Local food prices are therefore a good indicator of food scarcity and insecurity \citep{Brown2015}. \cite{Brown2015} investigate effects of weather disturbances and international price changes on local food prices which serve as a proxy for food scarcity. They use Kalman Filter approach (for details see \cite{KoopmanSJ}) and they focus on regions which contain large segments of low income population including Africa, South Asia and Latin America. They conclude that almost $20\%$ of local market prices are affected by domestic weather disturbances, $9\%$ of them are affected by international price change and $4\%$by both of them. Based on whether or not international food price and weather shocks are significant in explaining local food prices,\cite{brown2014book} groups food markets in developing countries into four categories: significantly affected by both international food prices and weather, significantly affected by weather but not international food prices, significantly affected by international food prices but not by weather and not significantly affected by either of them. \cite{brown2014book} then discus common characteristics of markets in each of these groups.

@@

\cite{Pedram2016} estimate national production losses per disaster worldwide during ${1964-2007}$ using a statistical method called superposed epoch analysis. Besides drought, they focus on extreme heat, cold and flood events. They conclude that on average $10.1\%$ reduction of cereal production can be linked to droughts and $9.1\%$ reduction is attributable to extreme heat. They did not find any significant effect of extreme cold and floods on production. \cite{Mehrabi2017} estimate cumulative crop production losses resulting from heat and drought disasters over the same time period (${1964-2007}$). Their estimates are almost half of those estimated by \cite{Pedram2016}. The biggest losses are in Botswana, Paraguay, Nigeria, Angola and USA.

\cite{OxfamIDS} uses the GLOBE Computable General Equilibrium model of the global economy to estimate food prices for various $2030$ scenarios. According to his results, climate change will cause $35\%$ decrease in maize productivity and $18\%$ decrease in productivity of paddy raise, wheat and other crops in south and south-east Africa. Climate change will also lead to substantial increase in both domestic and world market crop prices in comparison to baseline scenario in the absence of climate change. However, the increase in prices can be substantially mitigated if appropriate adaptation measures will be taken in sub-Saharan Africa \citet{OxfamIDS}.

>>>>

Shifts in staple food demand curve are usually not very large; hence, when staple food becomes scarce, its price is usually subject to a massive increase (\citealp{Brown2015, brown2014book}). For low income groups, this often leads to reduction in calorie intake and often to malnutrition and increased risk of related health problems (\citealp{Golden2011, Handa2006}). Local food prices are therefore a good indicator of food scarcity and insecurity \citep{Brown2015}. \cite{Brown2015} investigate effects of weather disturbances and international price changes on local food prices which serve as a proxy for food scarcity. They use Kalman Filter approach (see \cite{KoopmanSJ} for more details) and they focus on regions which contain large segments of low income population including locations in Africa, South Asia and Latin America. They conclude that almost $20\%$ of local market prices are affected by domestic weather disturbances, $9\%$ of them are affected by international price change and $4\%$by both of them. Based on whether or not international food price and weather shocks are significant in explaining local food prices, \cite{brown2014book} groups food markets in selected developing countries into four categories as follows: significantly affected by both international food prices and weather, significantly affected by weather but not international food prices, significantly affected by international food prices but not by weather and not significantly affected by either of them. \cite{brown2014book} then discus common characteristics of markets in each of these groups.

@@

\cite{Pedram2016} estimate national production losses per disaster worldwide during ${1964-2007}$ using a statistical method called superposed epoch analysis. Besides drought, they focus on extreme heat, cold and flood events. They conclude that on average $10.1\%$ reduction of cereal production can be linked to droughts and $9.1\%$ reduction is attributable to extreme heat. They did not find any significant effect of extreme cold and floods on production. \cite{Mehrabi2017} estimate cumulative crop production losses resulting from heat and drought disasters over the same time period (${1964-2007}$). Their estimates are almost half of those of \cite{Pedram2016}. The biggest losses are in Botswana, Paraguay, Nigeria, Angola and USA.

\cite{OxfamIDS} uses the GLOBE Computable General Equilibrium model of the global economy to estimate food prices for various $2030$ scenarios. According to his results, climate change will lead to substantial increase in both domestic and world market crop prices in comparison to baseline scenario in the absence of climate change. However, the increase in prices can be substantially mitigated if appropriate adaptation measures will be taken in sub-Saharan Africa \citep{OxfamIDS}.

### 2018-11-15 LiteratureFS/MyReview/Review12.tex
blll

\begin{document}
\sloppy
\section*{Definition of drought}
How we define drought is very important. Based on specification of drought, we can measure changes in aridity over time. According to international meteorological community, drought can be defined as \textit{ 'prolonged absence or marked deficiency of precipitation'}, a \textit{'deficiency of precipitation that results in water shortage for some activity or for some group'} or a \textit{'period of abnormally dry weather sufficiently prolonged for the lack of precipitation to cause a serious hydrological imbalance'} (\citealp{IPCCtrenberth, Heim2002}). The International Panel for Climate Change discuss three types of droughts: \textit{(i)}~'Agricultural drought' which is defined in terms of moister deficits in upper layer of soil up to about one meter depth~\textit{(ii)} 'meteorological drought' which refers to prolonged deficit of precipitation and~\textit{(iii)} 'hydrological drought' which relates to low streamflow, lake and levels of groundwater (IPCC; \citealp{IPCCtrenberth, Heim2002}). \cite{Trenberth2014} discuss definitions and measures of drought and their relation to contradictory results of two recent studies, in particular \cite{Sheffield2012} and \cite{Dai2011}. \cite{Sheffield2012} argue that drought has not increased much since $1960$ althought incorrect versions of Palmer Drought Severity Index (PDSI) give substantially different results. On the other hand, \cite{Dai2011} conclude that results differ only slightly for different forms of PDSI and all its forms indicates widespread drying. Besides difference in way of calculating the drought index, \cite{Trenberth2014} attribute the contradicting results to disparities among various rainfall datasets and different baseline periods.

As discussed by \cite{Trenberth2014}, drought can be measured in absolute terms (e.g. lake levels or amount of soil moisture) or using relative measures, such as PDSI. Because drought is defined based on one tail of probability distribution function of a drought measure, small decrease in mean can appear as very big increase in frequency of droughts. This has caused confusions and therefore usage of percentiles of soil moisture or streamflow is recommended as a better measure than mean \citep{Trenberth2014}.

Specifying a reliable index which could be used as a basis for definition of drought seems to problematic. The degree of drought does not only depend on precipitation, but also on whether and how fast the moisture is carried away (so the index should also incorporate evapotranspiration, which PDSI does. It also accounts for balance of precipitation.) Thus, besides precipitation, the index should incorporate humidity, wind, solar and long-wave radiation data \citep{vanderSchrier2011}.  However, availability of reliable data for solar radiation is a real problem \citep{Wang2012}.

\subsection*{How do authors define drought in terms of distribution of index?}
\textit{quantile..}
\begin{itemize}

\item \cite{Trenberth2014}:
\end{itemize}`

@@

Hazard can be defined as 'the potential occurrence of a natural or human induced
physical event that may cause loss of life, injury, or other
health impacts, as well as damage and loss to property, infrastructure,
livelihoods, service provision, and environmental resources' \citet{IPCC2012ch1}.
\section*{Changes and trends in drought}

@@

Shifts in staple food demand curve are usually not very large; hence, when staple food becomes scarce, its price is usually subject to a massive increase (\citealp{Brown2015, brown2014book}). For low income groups, this often leads to reduction in calorie intake and often to malnutrition and increased risk of related health problems (\citealp{Golden2011, Handa2006}). Local food prices are therefore a good indicator of food scarcity and insecurity \citep{Brown2015}. \cite{Brown2015} investigate effects of weather disturbances and international price changes on local food prices which serve as a proxy for food scarcity. They use Kalman Filter approach (see \cite{KoopmanSJ} for more details) and they focus on regions which contain large segments of low income population including locations in Africa, South Asia and Latin America. They conclude that almost $20\%$ of local market prices are affected by domestic weather disturbances, $9\%$ of them are affected by international price change and $4\%$by both of them. Based on whether or not international food price and weather shocks are significant in explaining local food prices, \cite{brown2014book} groups food markets in selected developing countries into four categories as follows: significantly affected by both international food prices and weather, significantly affected by weather but not international food prices, significantly affected by international food prices but not by weather and not significantly affected by either of them. \cite{brown2014book} then discus common characteristics of markets in each of these groups.

\cite{Ochieng2016} estimate effects of climate variability and change on agricultural production \footnote{Mesured as value of yields per acre in farm household} using panel data in Kenya. According to their results the effects are significant, yet different for different crops. Temperature has positive effect on tea and negative effect on production of maize and crop. Further, rainfall affects production of tea negatively.

@@

>>>>

\begin{document}
\sloppy
\section*{Definition of drought}
 According to international meteorological community, drought can be defined as \textit{ 'prolonged absence or marked deficiency of precipitation'}, a \textit{'deficiency of precipitation that results in water shortage for some activity or for some group'} or a \textit{'period of abnormally dry weather sufficiently prolonged for the lack of precipitation to cause a serious hydrological imbalance'} (\citealp{IPCCtrenberth, Heim2002}). The International Panel for Climate Change recognise three types of droughts: \textit{(i)}~'Agricultural drought' which is defined in terms of moister deficits in upper layer of soil up to about one meter depth~\textit{(ii)} 'meteorological drought' which refers to prolonged deficit of precipitation and~\textit{(iii)} 'hydrological drought' which relates to low streamflow, lake and levels of groundwater (IPCC; \citealp{IPCCtrenberth, Heim2002}).

 \cite{Trenberth2014} discuss different ways of formulation of the Palmer Drought Severity Index (\citealt{Sheffield2012, Dai2011}).  \cite{Sheffield2012} argue that way of formulation of PDSI can have substantial effect on estimation of changes in severity of droughts over time. According to \cite{Trenberth2014}, differences in formulation of the PDSI do not play a big role and it is more important to formulate the index such that the required data are available and reliable. \cite{Trenberth2014} attribute the differences in results of  \cite{Sheffield2012} and \cite{Trenberth2014} mostly to disparities among various rainfall datasets and different baseline periods rather than different formulations of the PDSI.

Drought can be measured in absolute terms (e.g. lake levels or amount of soil moisture) or in relative measures, which is for example the  PDSI \citep{Trenberth2014}. Because drought is defined based on one tail of probability distribution function of a drought measure, small decrease in mean can appear as substantial increase in frequency of droughts. This has caused confusions and therefore usage of percentiles of soil moisture or streamflow is recommended as a better measure than mean \citep{Trenberth2014}.

Specifying a reliable index which could be used as a basis for definition of drought seems to problematic. The degree of drought does not only depend on precipitation, but also on whether and how fast the moisture is carried away. Therefore, the index should also incorporate evapotranspiration, which PDSI does. Hence, \cite{vanderSchrier2011} argue that besides precipitation, the index should incorporate humidity, wind, solar and long-wave radiation data.   However, availability of reliable data for e.g. solar radiation is a real problem, hence, it may be preferable to use a more simplistic index (perhaps not including evapotranspiration) for which reliable data are available \citep{Wang2012}.

@@

Hazard can be defined as 'the potential occurrence of a natural or human induced
physical event that may cause loss of life, injury, or other
health impacts, as well as damage and loss to property, infrastructure,
livelihoods, service provision, and environmental resources' \citet{IPCC2012ch1}.

@@

Shifts in staple food demand curve are usually not very large. Hence, when staple food becomes scarce, its price is usually subject to a massive increase (\citealp{Brown2015, brown2014book}). For low income groups, this often leads to reduction in calorie intake, malnutrition and increased risk of related health problems (\citealp{Golden2011, Handa2006}). Local food prices are therefore a good indicator of food scarcity and insecurity (\citealp{baffes2017, Brown2015}). \cite{Brown2015} investigate effects of weather disturbances and international price changes on local food prices which serve as a proxy for food scarcity. They use Kalman Filter approach (see \cite{KoopmanSJ} for more details) and they focus on regions which contain large segments of low income population including locations in Africa, South Asia and Latin America. They conclude that almost $20\%$ of local market prices are affected by domestic weather disturbances, $9\%$ of them are affected by international price change and $4\%$by both of them. Based on whether or not international food price and weather shocks are significant in explaining local food prices, \cite{brown2014book} groups food markets in selected developing countries into four categories as follows: significantly affected by both international food prices and weather, significantly affected by weather but not international food prices, significantly affected by international food prices but not by weather and not significantly affected by either of them. \cite{brown2014book} then discus common characteristics of markets in each of these groups.

\cite{Ochieng2016} estimate effects of climate variability and change on agricultural production \footnote{Mesured as value of yields per acre in farm household} using panel data in Kenya. According to their results the effects are significant, yet different for different crops. Temperature has positive effect on tea and negative effect on production of maize and crop. Further, rainfall affects production of tea negatively. Another study, which finds positive correlation between precipitation and agricultural productivity is \cite{Vrieling2011}.

@@

\section*{Trends}

### 2018-11-15 LiteratureFS/MyReview/Review13.tex
blll

 \cite{Trenberth2014} discuss different ways of formulation of the Palmer Drought Severity Index (\citealt{Sheffield2012, Dai2011}).  \cite{Sheffield2012} argue that way of formulation of PDSI can have substantial effect on estimation of changes in severity of droughts over time. According to \cite{Trenberth2014}, differences in formulation of the PDSI do not play a big role and it is more important to formulate the index such that the required data are available and reliable. \cite{Trenberth2014} attribute the differences in results of  \cite{Sheffield2012} and \cite{Trenberth2014} mostly to disparities among various rainfall datasets and different baseline periods rather than different formulations of the PDSI.

@@

Specifying a reliable index which could be used as a basis for definition of drought seems to problematic. The degree of drought does not only depend on precipitation, but also on whether and how fast the moisture is carried away. Therefore, the index should also incorporate evapotranspiration, which PDSI does. Hence, \cite{vanderSchrier2011} argue that besides precipitation, the index should incorporate humidity, wind, solar and long-wave radiation data.   However, availability of reliable data for e.g. solar radiation is a real problem, hence, it may be preferable to use a more simplistic index (perhaps not including evapotranspiration) for which reliable data are available \citep{Wang2012}.

\subsubsection*{Extreme events, disaster and hazards}
\cite{IPCC2012ch1} define extreme events as 'the occurrence of a value of a weather or climate variable above (or below) a threshold value near the upper (or lower) ends of the range of observed values of the variable'. Some authors define extreme events only in terms of meteorological phenomena (\citealp{easterling2000, Jentsch2007}), others include also consequential physiological impacts or other effects on humans and ecosystems (\citealp{IPCC2012ch1, young2002}).

@@

\textcolor{blue}{look at many references in \cite{Brown2015} and \cite{Ochieng2016} !!!maybe also look if good references in \cite{OxfamIDS}??}

@@

>>>>

 \cite{Trenberth2014} discuss different ways of formulation of the Palmer Drought Severity Index (\citealt{Sheffield2012, Dai2011}).  \cite{Sheffield2012} argue that way of formulation of PDSI can have substantial effect on estimation of changes in severity of droughts over time. According to other studies, differences in formulation of the PDSI do not play a big role and it is more important to formulate the index such that the required data are available and reliable (\citealp{Trenberth2014,vanderSchrier2011,Wang2012}). \cite{Trenberth2014} attribute the differences in results of  \cite{Sheffield2012} and \cite{Trenberth2014} mostly to disparities among various rainfall datasets and different baseline periods rather than different formulations of the PDSI.

@@

%\textcolor{green}{Precist HEIM A Review of Twentieth Century Drought Indices Used in the United States 2002}
\subsubsection*{Extreme events, disaster and hazards}
\cite{IPCC2012ch1} define extreme events as 'the occurrence of a value of a weather or climate variable above (or below) a threshold value near the upper (or lower) ends of the range of observed values of the variable'. Some authors define extreme events only in terms of meteorological phenomena (\citealp{easterling2000, Jentsch2007}), others include also consequential physiological impacts or other effects on humans and ecosystems (\citealp{IPCC2012ch1, young2002}).

@@

%\textcolor{blue}{look at many references in \cite{Brown2015} and \cite{Ochieng2016} !!!maybe also look if good references in \cite{OxfamIDS}??}

@@

Large literature uses Ricardian approach to estimate impacts of climate change, including change in precipitation on crop revenue (\citealp{RicardianBello,kabubo2007, KMendelsohn2008, SeoMendelsohn}.

### 2018-11-15 LiteratureFS/MyReview/Review14.tex
blll

@@

Large literature uses Ricardian approach to estimate impacts of climate change, including change in precipitation on crop revenue (\citealp{RicardianBello,kabubo2007, KMendelsohn2008, SeoMendelsohn}.

>>>>

Large literature apply Ricardian approach (which was originally developed by \citealp{Mendelsohn94}) to estimate impacts of climate change, including change in precipitation on crop revenues in Africa (\citealp{RicardianBello,kabubo2007, KMendelsohn2008, SeoMendelsohn}). According to \cite{SeoMendelsohn} and \cite{KMendelsohn2008}, if the warming will be mild and wet crop net revenue should increase. In case of more severe and dry scenario, the crop revenue is likely to decrease (\citealp{KMendelsohn2008, SeoMendelsohn}).  \cite{RicardianBello} conclude that increase in temperature will affect net crop revenue negatively while increase in precipitation is likely to increase crop revenue. \citet{kabubo2007} argues that global warming will be harmful for crop production.

\cite{Deschenes2007Ric} use a production function approach to estimate economic impact of climate change, including impact of change in precipitation on agricultural profit in the United States. Production function method is an alternative to the Ricardian approach. \cite{Deschenes2007Ric} conclude that effect of global warming on agriculture should be positive in the US. In particular, agricultural yields should increase by $1.3$ billion US dollars or $4\%$. However, the estimated $95\%$ confidence interval is from $-0.5$ to $3.1$ billion US dollars, hence, the estimated effect is not very large.

@@

### 2018-11-15 LiteratureFS/MyReview/Review17.tex
blll

Numerous definitions of drought and their role are reviewed and discussed by \cite{wilhite1985} and \cite{wilhite2000}. They distinguish two main categories of definitions of drought: \textit{(i)} conceptual and \textit{(ii)} operational. Conceptual definitions are dictionary types, usually defining boundaries of the concept of drought\footnote{An example of conceptual definition of drought is an 'extended period - a season, a year, or several years of deficient rainfall relative to the statistical multi-year mean for a region' \cite{schneider1996}.}. Operational definitions are essential for an effective early warning system. An example of operational definition of agricultural drought can be obtaining the rate of soil water depletion based on precipitation and evapotranspiration rates and expressing these relationships in terms of drought effects on plant behaviour \citep{wilhite2000}.

\subsubsection*{Drought indices}
A numerical measure is needed to compare severity of drought across different time periods or geographical locations. However, as a result of a large disagreement about a definition of drought, there is no single universal drought index. Instead of that a number of measures of drought has been developed (\citealp{Heim2002, wilhite1985, wilhite2000}).

@@

The study of \cite{palmer1965} was a significant milestone in the history of quantification of drought severity. \cite{palmer1965} developed the Palmer Drought Severity Index (PDSI) using a complex water balance model. The PDSI is based on a hydrological accounting system, which incorporate antecedent precipitation, moisture supply and moisture demand (\citealp{Heim2002,palmer1965}). As the PDSI suffers from several weaknesses (for details see e.g. \citealt{Heim2002}), other indices were developed in the following decades. These include the standardized precipitation index (SPI) developed by \cite{SPI} and the standardized precipitation evapotranspiration index (SPEI) developed by \cite{SPEI}. The SPI specifies observed precipitation as a standardised departure from a chosen probability distribution which models the precipitation data. Values of SPI can be viewed as a multiple of standard deviations by which the observed amount of rainfall deviates from the long-term mean \citep{SPIonline}.\footnote{Can be created for various periods of 1-36 months, usually using monthly data.} The SPEI is similar to SPI, but unlike SPI, the SPEI includes the role of evapotranspiration (which captures increased temperature). It is based on water balance, therefore it can be compared to the self-calibrated PDSI \citep{SPEI}.

For an extensive overview of various drought indices see \cite{Heim2002}, \cite{monacelli2005}, \cite{svoboda2016} or \cite{zargar2011}. \cite{keyantash2002} quantify, evaluate and compare number of drought indices for meteorological, hydrological and agricultural forms of drought. Based on several criteria they conclude that rainfall deciles and SPI perform the best for meteorological drought.
%\cite{keyantash2002} further provides evaluation of the indecies...

\cite{Trenberth2014} discuss various formulations of the PDSI (\citealt{Sheffield2012, Dai2011}).  \cite{Sheffield2012} argue that the way  how PDSI is formulated can have substantial effect on estimation of changes in severity of droughts over time. According to other studies, differences in formulation of the PDSI do not play a big role and it is more important to formulate the index such that the required data are available and reliable (\citealp{Trenberth2014,vanderSchrier2011,Wang2012}). \cite{Trenberth2014} attribute the differences between the results of \cite{Sheffield2012} and \cite{Trenberth2014} to disparities among various rainfall datasets and to different baseline periods they use rather than to different formulations of the PDSI.

@@

Hazard can be defined as 'the potential occurrence of a natural or human induced
physical event that may cause loss of life, injury, or other
health impacts, as well as damage and loss to property, infrastructure,
livelihoods, service provision, and environmental resources' \citet{IPCC2012ch1}.
 \section*{Effects of Drought on Economy}

 \cite{Pedram2016} estimate national production losses per disaster worldwide during ${1964-2007}$ using superposed epoch analysis. Besides drought, they focus on extreme heat, cold and flood events. They conclude that on average $10.1\%$ reduction of cereal production can be linked to droughts and $9.1\%$ reduction is attributable to extreme heat. They did not find any significant effect of extreme cold and floods on production. \cite{Mehrabi2017} estimate cumulative crop production losses resulting from heat and drought disasters over the same time period (${1964-2007}$). Their estimates are almost half of those of \cite{Pedram2016}. The biggest losses are in Botswana, Paraguay, Nigeria, Angola and USA. \cite{HLAVINKA2009} analyse effects of drought on major cereal yields in the Czech Republic. They also define crop specific thresholds below which moisture deficit leads to severe negative impact on yields. \cite{HLAVINKA2009} find a link between severe droughts and reduction of crop yields.

 \cite{edwards2009} estimate economic and financial impact of drought in rural areas of Australia. They use data from a telephone interviews with $8000$ people. \cite{edwards2009} find that drought has significant effects on experience of financial hardship and detoriation of household with negative economic impact.

\cite{shahid2008} use SPI time series to produce drought hazard maps of western part of Bangladesh. The hazard maps are then combined with maps of socio-economic vulnerability to create drought risk maps.

@@

@@

\cite{robinson2010} use a Social Accounting Matrix (SAM) based CGE model to simulate impacts of food production and price shocks on household consumption in Ethiopia. They model a drought (or a disease) as a $20\%$ reduction in crop productivity and a $20\%$ decrease in livestock capital in lowland plains and drought-prone highlands in Ethiopia and they conclude that this would cause a real income loss on order of $2.3\%$ of Ethiopia's GDP. They also consider two alternative scenarios, i.e. variations in food aid and change in international prices.

\cite{OxfamIDS} uses the GLOBE Computable General Equilibrium model of the global economy to estimate food prices for various $2030$ scenarios. According to his results, climate change will lead to substantial increase in both domestic and world market crop prices in comparison to baseline scenario in the absence of climate change. However, the increase in prices can be substantially mitigated if appropriate adaptation measures will be taken in sub-Saharan Africa \citep{OxfamIDS}.
 \section*{Economic Effects of Climate and Weather}

\cite{Ochieng2016} estimate effects of climate variability and change on agricultural production\footnote{Measured as value of yields per acre in farm household.} using panel data in Kenya. According to their results, the effects are significant yet different for different crops. Temperature has positive effect on tea and negative effect on production of maize and crop. Furthermore, rainfall affects production of tea negatively. On the other hand, \cite{Vrieling2011} finds positive correlation between precipitation and agricultural productivity.

\subsubsection*{Hedonic approach and an alternative strategy}

\cite{Deschenes2007Ric} use an alternative strategy to estimate economic impact of climate change, including impact of change in precipitation, on agricultural profit in the United States.\footnote{The strategy of \cite{Deschenes2007Ric} is different from the hedonic approach and the production function approach.} In particular, they use year-to-year variation in precipitation and temperature which is assumed to be statistically independent of unobserved determinants of agricultural profit. \cite{Deschenes2007Ric} argue that this strategy is preferable to the hedonic approach, as hedonic approach is likely to be plagued with omitted variables bias. They conclude that effect of global warming on agriculture should be positive in the US. In particular, agricultural yields should increase by $1.3$ billion US dollars or $4\%$. However, the estimated $95\%$ confidence interval is from $-0.5$ to $3.1$ billion US dollars, hence, the estimated effect is not very large.

>>>>

\subsubsection*{Drought indices}
\textcolor{blue}{hello pridat referenci  sombreok et al 1982 viz \cite{kabubo2007}}
A numerical measure is needed to compare severity of drought across different time periods or geographical locations. However, as a result of a large disagreement about a definition of drought, there is no single universal drought index. Instead of that a number of measures of drought has been developed \citep{Heim2002}. For an extensive overview of various drought indices see for example  \cite{Heim2002} or \cite{keyantash2002}.

@@

The study of \cite{palmer1965} was a significant milestone in the history of quantification of drought severity. \cite{palmer1965} has developed the Palmer Drought Severity Index (PDSI) using a complex water balance model. The PDSI is based on a hydrological accounting system, which incorporate antecedent precipitation, moisture supply and moisture demand \citep{Heim2002}. As the PDSI suffers from several weaknesses (for details see e.g. \citealt{Heim2002}), other indices were developed in the following decades. These include the standardized precipitation index (SPI) developed by \cite{SPI} and the standardized precipitation evapotranspiration index (SPEI) developed by \cite{SPEI}. The SPI is based on cumulative probability of a given rainfall situation on a particular station. The SPEI is similar to SPI, but unlike SPI, the SPEI includes the role of temperature. It is based on water balance, therefore it can be compared to the self-calibrated PDSI \citep{SPEI}.

%\cite{keyantash2002} further provides evaluation of the indecies...

Drought can be measured in absolute terms (e.g. lake levels or amount of soil moisture) or in relative measures, which is for example the  PDSI \citep{Trenberth2014}. Because drought is defined based on one tail of probability distribution function of a drought measure, small decrease in mean can appear as a substantial increase in frequency of droughts. This has caused confusions and therefore usage of percentiles of soil moisture or streamflow is recommended as a better measure than mean \citep{Trenberth2014}.

\cite{Trenberth2014} discuss various formulations of the PDSI (\citealt{Sheffield2012, Dai2011}).  \cite{Sheffield2012} argue that the way  how PDSI is formulated can have substantial effect on estimation of changes in severity of droughts over time. According to other studies, differences in formulation of the PDSI do not play big role and it is more important to formulate the index such that the required data are available and reliable (\citealp{Trenberth2014,vanderSchrier2011,Wang2012}). \cite{Trenberth2014} attribute the differences between the results of \cite{Sheffield2012} and \cite{Trenberth2014} to disparities among various rainfall datasets and to different baseline periods they use rather than to different formulations of the PDSI.

@@

Hazard can be defined as 'the potential occurrence of a natural or human induced
physical event that may cause loss of life, injury, or other
health impacts, as well as damage and loss to property, infrastructure,
livelihoods, service provision, and environmental resources' \citet{IPCC2012ch1}.

 \section*{Economic Effects of Droughts}

\cite{Ochieng2016} estimate effects of climate variability and change on agricultural production\footnote{Mesured as value of yields per acre in farm household} using panel data in Kenya. According to their results, the effects are significant yet different for different crops. Temperature has positive effect on tea and negative effect on production of maize and crop. Furthermore, rainfall affects production of tea negatively. On the other hand, \cite{Vrieling2011} finds positive correlation between precipitation and agricultural productivity.

\cite{Pedram2016} estimate national production losses per disaster worldwide during ${1964-2007}$ using superposed epoch analysis. Besides drought, they focus on extreme heat, cold and flood events. They conclude that on average $10.1\%$ reduction of cereal production can be linked to droughts and $9.1\%$ reduction is attributable to extreme heat. They did not find any significant effect of extreme cold and floods on production. \cite{Mehrabi2017} estimate cumulative crop production losses resulting from heat and drought disasters over the same time period (${1964-2007}$). Their estimates are almost half of those of \cite{Pedram2016}. The biggest losses are in Botswana, Paraguay, Nigeria, Angola and USA.

@@

\subsubsection*{Ricardian and production function approach}

\cite{Deschenes2007Ric} use a production function approach (as an alternative to the Ricardian approach) to estimate economic impact of climate change, including impact of change in precipitation on agricultural profit in the United States. \cite{Deschenes2007Ric} conclude that effect of global warming on agriculture should be positive in the US. In particular, agricultural yields should increase by $1.3$ billion US dollars or $4\%$. However, the estimated $95\%$ confidence interval is from $-0.5$ to $3.1$ billion US dollars, hence, the estimated effect is not very large.

@@

\cite{robinson2010} use a Social Accounting Matrix (SAM) based CGE model to simulate impacts of food production and price shocks on household consumption in Ethiopia. They model a drought (or a disease) as a $20\%$ reduction in crop productivity and a $20\%$ decrease in livestock capital in lowland plains and drough-prone highlands in Ethiopia and they conclude that this would cause a real income loss on order of $2.3\%$ of Ethiopia's GDP. They also consider two alternative scenarios, i.e. variations in food aid and change in international prices.

\cite{OxfamIDS} uses the GLOBE Computable General Equilibrium model of the global economy to estimate food prices for various $2030$ scenarios. According to his results, climate change will lead to substantial increase in both domestic and world market crop prices in comparison to baseline scenario in the absence of climate change. However, the increase in prices can be substantially mitigated if appropriate adaptation measures will be taken in sub-Saharan Africa \citep{OxfamIDS}.
